---
title: "STAT 455: Assignment 4"
author: "Jo Dang"
output: pdf_document
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.height = 3, fig.width = 7)
```

One response to emergency department overcrowding in hospitals is “ambulance diversion”—closing its doors and forcing ambulances to bring patients to alternative hospitals. The California Office of Statewide Health Planning and Development collected data on how often hospitals enacted “diversion status”, enabling researchers to investigate factors associated with increasing amounts of ambulance diversions. The `ambulance3.csv` dataset contains observations from 184 California hospitals over a 3-year period (2013-2015):

Although there are other variables in the dataset, we will focus on the following variables. 


    - `diverthours` = number of hours of diversion status over the year     
    - `year2013` = years since 2013
    - `totalvisits1` = total number of patient visits to the emergency department over the year (in 1000s)
    - `ems_basic` = 1 if the emergency department can only handle a basic level of severity; 0 if the emergency department can handle higher levels of severity
    - `stations` = number of emergency department stations available for patients (fixed over 3 years)

```{r}
# Load needed packages
library(tidyverse)
library(lme4)
library(lmerTest)
library(Hmisc)
library(nlme)
summarize <- dplyr::summarize
```


```{r}
# read in data and select relevant variables
Hospitals <- read_csv("https://raw.githubusercontent.com/proback/BeyondMLR/master/data/ambulance3.csv")
Hospitals <- Hospitals %>% select(id, diverthours, year2013, totalvisits1, ems_basic, stations) %>% arrange(id, year2013)
head(Hospitals)
```

The full data are available at (https://github.com/proback/BeyondMLR/blob/master/data/ambulance3.csv). 

Note that most hospitals have three observations, one for each year in the dataset. 


## Question 1


### a) Explain why this can be thought of as a multilevel dataset. Which of the variables listed above are level one variables? Which are level two variables?  

There are individual / multiple observations on the same hospital, which means that the observations aren't all independent. So, the real observation unit is probably hostpitals. 

Level 1: diverthours, year2013
Level 2: id, totalvisits1, ems_basic, stations

### b) Create a histogram displaying the average number of divert hours per year for each hospital. Write a sentence describing the distribution.    

```{r}
ggplot(data=Hospitals,aes(x=diverthours)) + 
  geom_histogram(binwidth = 2, fill = "white",color = "black") + xlim(10,35) +
  xlab("Divert Hours") + ylab("Frequency") + labs(title="(b) Divert Hours") 
```
It seems that over the 3 years 2013-2015, many hospitals had about 17-19 hours of diversion. There are no signs of trends: there were similarly many hospitals with higher numbers of diversion hours as hospitals with lower numbers of diversion.

### c) Create a boxplot comparing EMS level to average number of divert hours. Write a sentence describing your findings.     

```{r}
ggplot(data=Hospitals,aes(x=factor(ems_basic), y=diverthours)) +
  geom_boxplot() + coord_flip() + ylab("EMS Level") + 
  xlab("Divert Hours") + labs(title="(c) EMS Level and Divert Hours")
```
Hospitals that can only handle a basic level of severity seem to frequently have more random outliers and higher diversion hours. Hospitals that can handle higher levels of severity seem to have more consistent trends of numbers of diversion hours.

### d) Create a scatterplot comparing number of stations to average number of divert hours. Write a sentence describing your findings.    

```{r}
ggplot(Hospitals, aes(stations,diverthours))+
  geom_point() + xlab("Numbers of Stations") + 
  ylab("Divert Hours") + labs(title="(c) Numbers of Stations and Divert Hours")
```
Although there is not a strict trend, hospitals with significantly higher numbers of stations (>=65) do have fewer hours of diversion. But, generally, the hospitals have similar hours of diversion despite having different numbers of stations.

### e) Create lattice plots displaying the number of divert hours over time for the 15 hospitals in the `Hospitals_sample` dataset given below. Write a sentence describing your findings. 

```{r}
Hospitals_Sample <- Hospitals %>% filter(id %in% c(106010735, 106190432, 106190017, 106190529, 106190125, 106190243, 106504042, 106190413, 106331164,  106190796,  106190687, 106301317, 106430883,  106190148))
```


```{r}
ggplot(Hospitals_Sample, aes(x=year2013,y=diverthours)) +
geom_point() + geom_smooth(method="lm",color="black") +
facet_wrap(~(id),ncol=5) +
theme(strip.text.x=element_blank()) + ylab("Divert Hours") + xlab("Years Since 2013") +
  labs(title="(e) Divert Hours Over The Years")
```
The general trends are the more time passed, the hours of diversion hospitals either increased or stayed the same, with a few exceptions when not enough data was available to show trends.

### f) Create latticed spaghetti plots that illustrate the relationship between divert hours and  EMS level. Include all hospitals, not only those in the sample above. Also include a trendline, using geom_smooth(). Write a sentence describing your findings.   

```{r}
ggplot(Hospitals, aes(x = id, y = diverthours)) + 
  geom_line(color="grey") + geom_smooth() +
  facet_wrap(~ems_basic) + xlab("Hospitals") + ylab("Divert Hours") +
  labs(title="(f) Divert Hours and EMS Levels")
```
Hospitals that can only handle basic ems level emergencies have fewer and less drastic fluctuations in their numbers of diversion hours.

### g) Create latticed spaghetti plots that illustrate the relationship between divert hours and number of stations, broken down by quartile. Include all hospitals, not only those in the sample above. Also include a trendline, using geom_smooth(). (Hint: Use the `cut2` function, which is part of the `Hmisc` package.) Write a sentence describing your findings.    


```{r}
ggplot(Hospitals, aes(x = id, y = diverthours)) + 
  geom_line(color="grey") + geom_smooth() +
  facet_wrap(~cut(stations,4)) + xlab("Hospitals") + ylab("Divert Hours") +
  labs(title="(g) Divert Hours and Stations")
```
Most of the hospitals in the data sets had numbers of stations within the 25-75 quartiles, and they all have different and fluctuating numbers of diverst hours. There doesn't seem to be a trend or correlation between number of station and divert hours based on these surface observations.

## Question 2

### a) Let $Y_{ij}$ represent the number of divert hours for the ith hospital in the jth year. Fit an unconditional means model, of the form

\[
Y _{ij} = \alpha_{0} + u_{i} + \epsilon_{ij} \textrm{ with } u_{i} \sim N(0, \sigma^2_u) \textrm{ and } \epsilon_{ij} \sim N(0, \sigma^2)
\]

### to the data. 

```{r}
uM <- lmer(diverthours ~ 1 + (1|id), 
                REML=T, data=Hospitals)
summary(uM)
```

### b) Calculate and interpret the intraclass correlation coefficient.  

The average correlation for any pair of responses from the same hospital is 0.6481.

### c) Fit an unconditional growth model of the form

\[
Y_{ij}=\alpha_{0} + \beta_{0}\textrm{Year2013}_{ij}+u_{i}+v_{i}\textrm{Year2013}_{ij} + \epsilon_{ij}
\]

### where $\epsilon_{ij}\sim N(0,\sigma^2)$ and

\[
 \left[ \begin{array}{c}
            u_{i} \\ v_{i}
          \end{array}  \right] \sim N \left( \left[
          \begin{array}{c}
            0 \\ 0
          \end{array} \right], \left[
          \begin{array}{cc}
            \sigma_{u}^{2} & \\
            \rho_{uv}\sigma_{u}\sigma_{u} & \sigma_{v}^{2}
          \end{array} \right] \right) . 
\]


```{r}
uG <- lmer(diverthours~ year2013 + (year2013|id), 
  REML=T, data=Hospitals)
summary(uG)
```


### d) For the model in the previous part, give the values of the estimates for parameters $\alpha_0$, $\beta_0$, $\sigma_u$, $\sigma_v$, $\sigma^2$, and $\rho_{uv}$. Interpret these estimates in context.   

$\alpha_0$ = 356.36 = the mean divert hours for the population of hospitals in 2013
$\beta_0$ = 140.14 = the mean yearly change in divert hours for the population during the three-year observation period
$\sigma_u$ = 562.0 = the standard deviation in divert hours between hospitals in 2013
$\sigma_v$ = 276.1 = the standard deviation in rates of change in divert hours during the three-year observation period
$\sigma^2$ = 112306 = the variability in divert hours within hospitals
$\rho_{uv}$ = 0.13 = the correlation in hospitals’ 2013 divert hours and their rate of change in divert hours between 2013 and 2015

### e) Fit a model that includes both year, and EMS level as fixed effects. Do not include an interaction term. Assume that that divert hours vary randomly between hospitals, and that the yearly rate of change of divert hours also varies between hospitals. Display the summary output. You do not need to interpret model coefficients.

```{r}
m1 <- lmer(diverthours ~ year2013 + ems_basic +(year2013|id),REML=T,data=Hospitals)
summary(m1)
```


### f) Modify the model in (e), to assume that divert hours vary randomly between hospitals, but change at a yearly rate that is the same for all hospitals. Display the summary output. You do not need to interpret model coefficients.

```{r}
m2 <- lmer(diverthours ~ year2013 + ems_basic +(1|id),REML=T,data=Hospitals)
summary(m2)
```


### g) Use AIC and BIC to compare the models in (e) and (f). Which model is preferred? What does this say about the relationship between divert hours at hospitals over time?

```{r}
AIC(m1,m2)
```

```{r}
BIC(m1,m2)
```

m1 is preferred as suggested by both the AIC and BIC. This means that there likely are differences in rates of change in divert hours throughout the years between hospitals.

### h) Add an interaction term relating year and EMS level to the preferred model in (g). Interpret the interaction coefficient in context.    

```{r}
m3 <- lmer(diverthours ~ year2013 + ems_basic + year2013:ems_basic +(year2013|id),REML=T,data=Hospitals)
summary(m3)
```

Hospitals with only a basic level of ems treatments have an estimated mean decrease in divert hours of 944.2 hours per year over the three-year observation period, 266.1 hours more than the mean yearly decrease among hospitals that can treats more than basic ems treatments.

### i) Add stations to the model in (h). Also add an interaction term between stations and year. Use a likelihood ratio test to compare this model to the model in (h). Interpret your results. Does number of stations at a hospital appear to be related to divert hours? Is there evidence of an interaction between year and number of stations? If so, what should we conclude from this?  

```{r}
m4 <- lmer(diverthours ~ year2013 + ems_basic + stations + year2013:ems_basic + year2013:stations + (year2013|id),REML=T,data=Hospitals)
summary(m4)
```

```{r}
likelihood <- anova(m3, m4, test = "Chisq")
likelihood
```
The small p-value provides evidence against the null hypothesis that a possible relationship is purely based on chance, or, in other words, that model 3 is sufficient. This suggests that the number of stations at a hospital does affect divert hours and there is an interaction between year and number of stations. Then, model 4 would be a better final model.

## Question 3

### a) Using the same fixed and random effects from your preferred model so far, fit a first order autoregressive (AR-1) time series model. Use the code below to filter the dataset to hospitals that have data for all 3 years, and fit the model using only these hospitals. Display a summary of the model output. 

```{r}
# keep only hospitals with observed scores for all 3 years
Hospitals <- Hospitals %>% select(-totalvisits1)
Hospitals.wide <- pivot_wider(Hospitals, names_from = "year2013", values_from = "diverthours")
Hospitals_all3yrs <- Hospitals %>% filter(id %in% Hospitals.wide[complete.cases(Hospitals.wide),]$id)
```

```{r}
AR1.sim = gls(diverthours ~  year2013 + ems_basic + stations + year2013:ems_basic + year2013:stations, data = Hospitals,
    correlation = corAR1(form = ~ year2013|id))

summary(AR1.sim)
```

### b) Report the estimate of $\phi$ and interpret it in context.   

The correlation in divert hours at the same hospital, one year apart is estimated to be $\phi$ = 0.72.
